Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-14 paper/paper.tex
sum eta law

\subsubsection{Lambek Calculus and Categorial Grammar}
Lambek's original syntactic calculus \cite{lambek58} describes a logical system
for linguistic derivations. Later, he explicitly gives semantics to his
syntactic calculus inside of a non-commutative biclosed monoidal category
\cite{lambek1988categorial}. This led to many uses of non-commutative linear logic and lambda calculi in
linguistics \cite{buszkowskiTypeLogicsGrammar2003}. Grammars based on logic in
this manner may go by the names Lambek calculus, typelogical grammar, or
categorial grammar. Such categorial grammars correspond to the subtheory of our
calculus generated by $\otimes$, $\lto$, and $\tol$.

@@

\subsubsection{Typed Grammars}
The usage of a simple type system to reason about regular expressions up to weak
equivalence was
introduced by Frisch and Cardelli \cite{frischCardelli}. Hengelin and Nielsen
\cite{henglein_regular_2011} later adapt this approach by providing a type
system for regular expressions whose semantics were in one-to-one correspondence
with the set of parse trees.
Elliott \cite{elliottSymbolicAutomaticDifferentiation2021} gives a similar
semantic interpretation of regular grammars as type-level predicates on strings. The type system of \theoryabbv extends these semantic interpretations of to a broader class of grammars and also gives a formal syntax and equational theory for the parse transformers that these prior works lack.

@@

\subsection{Type Checking and Elaboration}
Our focus in this work has been on the verification of parsers for
grammars over strings, but because \theoryabbv allows for the
definition of arbitrarily powerful grammars, the system could also be
used in principle for more sophisticated semantic analysis such as
scope checking or type checking. Alternatively, we could more directly
encode type type systems as linear types in a modified version of
\theoryabbv where linear types are not grammars over strings, but
\emph{type systems} over trees. This could analogously serve as a
framework for verified type checking and static analysis.

@@

\paragraph{$\bigoplus\eta$}
It suffices to show
\[
  \semg{\letin {\sigma~x~a} {c'} {f[\sigma~x~a/c]}} = \semg{f[c'/c]}
\]
\begin{proof}
  \begin{align*}
    \semg{\letin {\sigma~x~a} {c'} {f[\sigma~x~a/c]}}
    &= [ \sem{f[\sigma~x~a/c]} ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m
  \end{align*}
\end{proof}
\max{todo: finish}

>>>>

%% \subsubsection{Lambek Calculus and Categorial Grammar}
Lambek's original syntactic calculus \cite{lambek58} describes a logical system
for linguistic derivations. Later, he explicitly gives semantics to his
syntactic calculus inside of a non-commutative biclosed monoidal category
\cite{lambek1988categorial}. This led to many uses of non-commutative linear logic and lambda calculi in
linguistics \cite{buszkowskiTypeLogicsGrammar2003}. Grammars based on logic in
this manner may go by the names Lambek calculus, typelogical grammar, or
categorial grammar. Such categorial grammars correspond to the subtheory of our
calculus generated by $\otimes$, $\lto$, and $\tol$.

@@

%% \subsubsection{Typed Grammars}
The usage of a simple type system to reason about regular expressions up to weak
equivalence was
introduced by Frisch and Cardelli \cite{frischCardelli}. Hengelin and Nielsen
\cite{henglein_regular_2011} later adapt this approach by providing a type
system for regular expressions whose semantics were in one-to-one correspondence
with the set of parse trees.
Elliott \cite{elliottSymbolicAutomaticDifferentiation2021} gives a similar
semantic interpretation of regular grammars as type-level predicates on strings. The type system of \theoryabbv extends these semantic interpretations of to a broader class of grammars and also gives a formal syntax and equational theory for the parse transformers that these prior works lack.

@@

\subsection{Type Checking and Semantic Analysis}
Our focus in this work has been on the verification of parsers for
grammars over strings, but because \theoryabbv allows for the
definition of arbitrarily powerful grammars, the system could also be
used in principle for more sophisticated semantic analysis such as
scope checking or type checking. Alternatively, we could more directly
encode type type systems as linear types in a modified version of
\theoryabbv where linear types are not grammars over strings, but
\emph{type systems} over trees. This could analogously serve as a
framework for verified type checking and static analysis.

@@

\paragraph{$\bigoplus\eta$}
It suffices to show
\[
  \semg{\letin {\sigma~x~a} {c'} {f[\sigma~x~a/c]}} = \semg{f[c'/c]} = \semg{f}
  \]
First, expanding the left hand side, we have.
\begin{proof}
  \begin{align*}
    \semg{\letin {\sigma~x~a} {c'} {f[\sigma~x~a/c]}}
    &= [ \semg{f[\sigma~x~a/c]} ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
    &= [ \semg{f} \circ (id \otimes i_x) \otimes id ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
    &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m\\
  \end{align*}
Since the domain has the universal property of a coproduct (due to distributivity), to prove
this is equal to $\semg{f}$, it is sufficient to prove they are equal
when composed with the injections:

\begin{align*}
  \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \circ (id \otimes i_y) \otimes id
  &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ d  \circ (id \otimes i_y) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \tag{naturality}\\
  &= \semg{f} \circ [ (id \otimes i_x) \otimes id \circ (m^{-1})]_{(x : X)} \circ i_y \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \tag{naturality}\\
  &= \semg{f} \circ (id \otimes i_y) \otimes id \circ (m^{-1}) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \\
  &= \semg{f} \circ (m^{-1}) \otimes id \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m \circ (id \otimes i_y) \otimes id \\
  &= \semg{f} \circ (id \otimes i_y) \otimes id \tag{coherence}\\
\end{align*}
\end{proof}
